---
title: "Mapping Report"
fig_width: 6
output:
  word_document: default
  html_document: default
fig_height: 4
---

```{r include=FALSE}
library(tidyverse)
library(lctools)
library(RColorBrewer)
```

# Importing Data

The `lctools` package provides The `GR.Municipalities` data. Which is a `SpatialPolygonDataFrame` object. The `data` slot contains the demographic information of the 325 municipalities of Greece.

```{r}
data(GR.Municipalities)
data_by_Municipality <- GR.Municipalities@data
```

# Classes creation for Unemployment Rate

We will use `cut` function to create 6 classes of unemployment rate.

```{r}
k = 6
data_by_Municipality$UnemrT_cluster <- cut(data_by_Municipality$UnemrT01, breaks = k)
table(data_by_Municipality$UnemrT_cluster)
```

However, the classes breaks provided by the `cut` function are not rounded to integers. We will use some manipulations to round breaks to the nearest integer.

```{r}
classes_bandwidth <- group_by(data_by_Municipality, UnemrT_cluster) %>% 
    summarise(min_ban = round(min(UnemrT01)),
              max_ban = round(max(UnemrT01)))

classes_bandwidth <- c(classes_bandwidth$min_ban, classes_bandwidth$max_ban[k])
classes_bandwidth
```

Next, we construct the labels for the classes in the form : $[x,x+1[$ 

```{r}
class_labels <- paste0("[", classes_bandwidth, ",", lead(classes_bandwidth), "[")
class_labels <- class_labels[-(k+1)]
class_labels
```


We will assign municipalities to the new classes according to the new breaks we have constructed.

```{r}
#Function serving to assign unemployment rate to corresponding class
cluster_fun <- function(x, Bands, Labels) {
  k = length(Bands)
  for (i in 2:k) {
    if (x >= Bands[i-1] & x < Bands[i])
      return(Labels[i-1])
  }
}

data_by_Municipality <- rowwise(data_by_Municipality) %>% 
  mutate(class_unemr = cluster_fun(UnemrT01, classes_bandwidth, class_labels),
         class_unemr = factor(class_unemr, levels = class_labels))

head(data_by_Municipality[c(1:4,11,16)])

```

# Exploring Unemployment Rate classes

Before going to map creation, let's explore the distribution of unemployment rates within classes.

```{r}
#Custom theme to use for coming plots
Custom_theme <- theme_bw()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "None")
```

First, let's visualize how many municipalites in each group using bar plot
```{r fig.align = 'center'}
table(data_by_Municipality$class_unemr)

ggplot(data_by_Municipality, aes(x = class_unemr, fill = class_unemr))+
  geom_bar()+
  Custom_theme+
  scale_fill_brewer(palette = "Greens")+
  labs(x = "Classes of Unemployment Rate",
       y = "Number of Municipalities")
```

More than half of municipalities have unemployment rates between 4% and 7%. Fewer municipalites have unemployment rates between 13% and 20%.

Boxplots help visualising distribution of unemployment rate within each class.

```{r fig.align = 'center'}

ggplot(data_by_Municipality, aes(fill = class_unemr))+
  geom_boxplot(aes(y = UnemrT01))+
  scale_y_continuous(breaks = 0:20)+
  facet_grid(~class_unemr, scale = "free")+
  Custom_theme +
  scale_fill_brewer(palette = "Greens")+
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank())+
  labs(y = "Unemployment Rate")
```


# Data preparation for mapping

Preparing data for mapping includes assigning polygons in the `sp` object to our data. So that each municipality is represented with a finite number of segments connected to form a closed polygon
.
```{r}
map_data <- fortify(GR.Municipalities, region = "OBJECTID") %>% 
  mutate(id = as.numeric(id)) %>%
  left_join(data_by_Municipality, by = c("id" = "OBJECTID"))
```


# Mapping

```{r fig.height= 10, fig.width= 14, fig.align = 'center'}
map_data %>% 
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(color = "gray80", aes(fill = class_unemr)) +
  scale_fill_brewer(palette = "Greens") +
  coord_equal() +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(x= "", y= "", fill= "Class of unemployment rate")
```

